The forestecology R package for fitting and assessing models of interspecies competition on the growth of trees

### 2021-02-22 etc/paper/forestecology.Rmd
Rewrite "permutation test" and "spatial cross-validation" sections

---
title: "The `forestecology` R package for modeling interspecies competition between trees"

@@

authors: 
- name: Albert Y. Kim 
  thanks: "Albert Y. Kim is Assistant Professor, Statistical & Data Sciences, Smith College, Northampton, MA 01063 (e-mail: akim04@smith.edu)."
  affiliation: Program in Statistical & Data Sciences, Smith College

@@

abstract: 
- "Move abstract below here after completed."

keywords:
- forest ecology, competition, R, Rstats, tidyverse, sf, cross-validation,

@@

# Abstract (350 words) {-}

1. When modeling growth of trees forest ecologists often incorporate the effect of interspecies competition. Many such models are based on a neighborhood effect assumption whereby all trees within a fixed distance of all focal trees are considered competitors. Methods are needed to evaluate the effect of interspecies competition and to assess their quality. 
1. We present the `forestecology` package providing methods for both 1) evaluating the out-of-sample performance of our model using spatial-crossvaliation and 2) testing a null hypothesis of of no impact of competitor species' identity on the growth of trees using a permutation test. We implement a class and methods using R's S3 object-oriented system, for a specific linear, Bayesian neighborhood competition model of tree growth. 
1. We demonstrate the package's functions using data from the Smithsonian Conservation Biology Institute's large forest dynamics plot, part of the ForestGEO network of reseach sites. Given ForestGEO's data collection protocols and data formatting standards, the package cross-compatibility of code. We show both that 1) competitor species identity matters and 2) that not spatially cross-validating leads to error estimates that are overly optimistic. 
1. The package follows `tidyverse`-like structure whereby verb-named functions can be modularly "piped" in sequence to intuitively display the sequence of steps of analysis from start to finish. Additionally, most inputs/outputs of functions assume an are of `sf` class from the simple features package, thereby facilitating all wrangling and visualization of geospatial data. Lastly, even though our package is currently limited to one specific model, the package is setup such that it can be easily extended to other models.

@@

1. Compute the growth of trees based on census data.
1. Add spatial information:
    1. Define buffer region trees.
    1. Add spatial cross-validation block information.
1. Identify all focal and corresponding competitor trees.
1. Fit model and make predictions.

@@

## Compute the growth of trees based on census data

@@

```{r scbi-trees, out.width="66%", fig.cap="Compute growth of trees based on census data from the Smithsonian Conservation Biology Institute (SCBI)."}
ggplot() +
  geom_sf(data = growth_scbi, aes(size = growth)) + 
  scale_size(breaks = c(0.01, 0.1, 1), range = c(0.1, 1))
```

## Add spatial information {#spatial-information}

@@

```{r scbi-spatial-information, out.width="66%", fig.cap="Add spatial information: Buffer region and spatial cross-validation blocks (1 through 4). All trees in the interior of the study region will be the focal trees whose growth will modeled."}
ggplot() +
  geom_sf_text(data = growth_scbi, aes(label = foldID, col = buffer)) +
  geom_sf(data = blocks_scbi, fill = "transparent")
```

## Identify all focal and corresponding competitor trees {#focal-vs-comp}

@@

Using the `unnest()` function from the `tidyr` package, we can flatten list-column into regular columns. We observe that for the same focal tree with DBH equal to 13.65cm, we have information on all 20 competitor trees whose `dist` distance to the focal tree is less than or equal to `r comp_dist`, including their unique tree-stem ID number, their species, and their basal area (in m$^2$) calculated as $\frac{\pi \times (\text{DBH/2})^2}{10000}$ where $DBH$ is the value from the earlier census in cm. Saving our focal versus competitor information in list-column minimizes redundancy since we do not repeat information on the focal tree 20 times.

```{r, eval = FALSE}
focal_vs_comp_scbi %>% 
  filter(focal_ID == 4) %>% 
  select(focal_ID, dbh, comp) %>% 
  unnest(cols = "comp")
```
```{r scbi-focal-vs-comp-3, echo = FALSE}
# Not sure why we need to do things this way:
focal_vs_comp_scbi %>% 
  filter(focal_ID == 4) %>% 
  select(focal_ID, dbh, comp) %>% 
  unnest(cols = "comp") %>% 
  print(n = 5)
```
```{r scbi-focal-vs-comp-map, out.width="66%", echo = FALSE, fig.cap="Identify all focal and corresponding competitor trees: All 20 competitor trees of focal tree 4."}
comp_radius <- growth_scbi %>% 
  filter(stemID == 4) %>% 
  st_buffer(dist = comp_dist)

@@

## Fit model and make predictions {#model-fit-predict}

@@

Lastly, the generic for `ggplot2::autoplot()` allows us to

```{r scbi-posterior-viz, out.width="100%", fig.cap="Fit model: Posterior distributions of all parameters for three species."}
# Plot posteriors for only a subset of species
sp_to_plot <- c("litu", "quru", "cagl")

@@

HEY BERT PICK IT UP HERE

```{r scbi-permutation-model-fit, eval = FALSE}
comp_bayes_lm_scbi <- focal_vs_comp_scbi %>%
  comp_bayes_lm(prior_param = NULL, run_shuffle = TRUE)

focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  mutate(
    growth_hat = predict(comp_bayes_lm_scbi, newdata = focal_vs_comp_scbi)
  )

model_rmse <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
model_rmse
```

@@

TODO: Do we talk about `add_buffer_variable(direction = "out")` when each fold gets its turn being the training data?

The model fits and predictions in Section \ref{model-fit-predict} all suffer from a common failing: they use the same data to both fit the model and to assess the model's performance using the RMSE. As argued by @roberts_cross-validation_2017, this can lead to overly optimistic assessments of model quality as the models can be overfit, in particular in situations where spatial-autocorrelation is present. To mitigate the effects of such overfitting, we use a spatially block cross-validation algorithm implemented in the `run_cv()`. This function at its core uses the same model fitting implemented in the `fit_bayesian_model()` function, however trains the model on $k-1$ spatial folds of the train and returns fitted values for the test data. Recall that the spatial blocking scheme wass encoded in Section \ref{spatial-information}.

@@

```{r}
focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
```

@@

>>>>

---
title: "The forestecology R package for fitting and assessing models of interspecies competition on the growth of trees"

@@

authors: 
- name: Albert Y. Kim 
  affiliation: Program in Statistical & Data Sciences, Smith College

@@

abstract: 
- |
  | 1. When modeling growth of trees forest ecologists often incorporate the effect of interspecies competition. Many such models are based on a neighborhood effect assumption whereby all trees within a fixed distance of all focal trees are considered competitors. Methods are needed to evaluate the effect of interspecies competition and to assess their quality. 
  | 2. We present the `forestecology` package providing methods for both 1) evaluating the out-of-sample performance of our model using spatial-crossvaliation and 2) testing a null hypothesis of of no impact of competitor species' identity on the growth of trees using a permutation test. We implement a class and methods using R's S3 object-oriented system, for a specific linear, Bayesian neighborhood competition model of tree growth. 
  | 3. We demonstrate the package's functions using data from the Smithsonian Conservation Biology Institute's large forest dynamics plot, part of the ForestGEO network of reseach sites. Given ForestGEO's data collection protocols and data formatting standards, the package cross-compatibility of code. We show both that 1) competitor species identity matters and 2) that not spatially cross-validating leads to error estimates that are overly optimistic. 
  | 4. The package follows `tidyverse`-like structure whereby verb-named functions can be modularly "piped" in sequence to intuitively display the sequence of steps of analysis from start to finish. Additionally, most inputs/outputs of functions assume an are of `sf` class from the simple features package, thereby facilitating all wrangling and visualization of geospatial data. Lastly, even though our package is currently limited to one specific model, the package is setup such that it can be easily extended to other models.

keywords:
- forest ecology, competition, R, Rstats, tidyverse, sf, cross-validation, spatial statistics

@@

@@

- Step 1: Compute the growth of trees based on census data.
- Step 2: Add spatial information:
    1. Define buffer region trees.
    1. Add spatial cross-validation block information.
- Step 3: Identify all focal and corresponding competitor trees.
- Step 4: Fit model, as well as:
    1. Compute fitted/predicted values.
    1. Visualize posterior distributions.

@@

## Step 1: Compute the growth of trees based on census data

@@

```{r scbi-trees, out.width="66%", fig.cap="Step 1 - Compute growth of trees based on census data: All trees from a 9 ha subsection of the Smithsonian Conservation Biology Institute (SCBI) forest plot."}
ggplot() +
  geom_sf(data = growth_scbi, aes(size = growth)) + 
  scale_size(breaks = c(0.01, 0.1, 1), range = c(0.1, 1))
```

## Step 2: Add spatial information {#spatial-information}

@@

```{r scbi-spatial-information, out.width="66%", fig.cap="Step 2 - Add spatial information: Buffer region and spatial cross-validation blocks (1 through 4). All trees in the interior of the study region will be the focal trees whose growth will modeled."}
ggplot() +
  geom_sf(data = blocks_scbi, fill = "transparent", linetype = "dashed") +
  geom_sf_text(data = growth_scbi %>% sample_n(1000), aes(label = foldID, col = buffer))
```

## Step 3: Identify all focal and corresponding competitor trees {#focal-vs-comp}

@@

Using the `unnest()` function from the `tidyr` package, we can flatten list-column into regular columns. We observe that for the same focal tree with DBH equal to 13.65cm, we have information on all 20 competitor trees whose `dist` distance to the focal tree is less than or equal to `r comp_dist`, including their unique tree-stem ID number, their species, and their basal area (in m$^2$) calculated as $\frac{\pi \times (\text{DBH/2})^2}{10000}$ where $DBH$ is the value from the earlier census in cm. Saving our focal versus competitor information in list-column minimizes redundancy since we do not repeat information on the focal tree 20 times. The spatial distribution of these trees is visualized in Figure \ref{fig:scbi-focal-vs-comp-map}: the dashed circle extends `r comp_dist` m away from the focal tree while all 20 competitor trees are within this circle.

```{r, eval = FALSE}
focal_vs_comp_scbi %>% 
  filter(focal_ID == 4) %>% 
  select(focal_ID, dbh, comp) %>% 
  unnest(cols = "comp")
```
```{r scbi-focal-vs-comp-3, echo = FALSE}
# Not sure why we need to do things this way:
focal_vs_comp_scbi %>% 
  filter(focal_ID == 4) %>% 
  select(focal_ID, dbh, comp) %>% 
  unnest(cols = "comp") %>% 
  print(n = 5)
```
```{r scbi-focal-vs-comp-map, out.width="66%", echo = FALSE, fig.cap="Step 3 - Identify all focal and corresponding competitor trees: All 20 competitor trees of focal tree 4."}
comp_radius <- growth_scbi %>% 
  filter(stemID == 4) %>% 
  st_buffer(dist = comp_dist)

@@

## Step 4: Fit model {#model-fit-predict}

@@

Lastly, the generic for `ggplot2::autoplot()` allows us to plot the posterior distribution of all parameters in Figure \ref{fig:scbi-posterior-viz}.

```{r scbi-posterior-viz, out.width="100%", fig.cap= "Step 4 - Fit model: Posterior distributions of all parameters for three species."}
# Plot posteriors for only a subset of species
sp_to_plot <- c("litu", "quru", "cagl")

@@

We use the four steps of our analysis sequence we answer the first of our two questions: does competitor species identity matter when modeling the growth of focal trees? We answer this via a permutation test: Under a null hypothesis where competitor species identity does not matter we can permute this variable within each focal tree, compute the RMSE (the test statistic of interest), repeat this process several times to construct a null distribution of RMSE, and compare the observed RMSE to assess significance. Going back to our example in Section \ref{focal-vs-comp} inspecting all 20 competitor trees to focal tree with `focal_ID` 4, the permutation test involves randomly resampling the `comp_sp` variable with replacement, leaving all other competitor variables intact. The resampling with replacement is nested within each focal tree in order to preserve the neighborhood structure of our competitor model. To run the permutation test, we use the same code as in Section \ref{model-fit-predict}, but adding a `run_shuffle = TRUE` argument to `comp_bayes_lm()`.

```{r scbi-permutation-model-fit}
comp_bayes_lm_scbi_shuffle <- focal_vs_comp_scbi %>%
  comp_bayes_lm(prior_param = NULL, run_shuffle = TRUE)

focal_vs_comp_scbi <- focal_vs_comp_scbi %>%
  mutate(growth_hat_shuffle = predict(comp_bayes_lm_scbi_shuffle, newdata = focal_vs_comp_scbi))

model_rmse_shuffle <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat_shuffle) %>%
  pull(.estimate)
model_rmse_shuffle
```

The resulting RMSE of `r model_rmse_shuffle` based on the permutation test is `r ifelse(model_rmse_shuffle < model_rmse, "smaller", "larger")` than the earlier RMSE of `r model_rmse`, suggesting that models that do incorporate competitor species identity better fit the data. We conduct a fuller simulation in Section below.

@@

Using the four steps of our analysis sequence again, we answer the second of our two questions: how can we obtain an accurate estimate of model performance/error?  The model fits and predictions in Section \ref{model-fit-predict} all suffer from a common failing: they use the same data to both fit the model and to assess the model's performance using the RMSE. As argued by @roberts_cross-validation_2017, this can lead to overly optimistic assessments of model quality as the models can be overfit, in particular in situations where spatial-autocorrelation is present. To mitigate the effects of such overfitting, we use a spatially block cross-validation algorithm.

To this end, we use the `foldID` variable defined in Section \ref{spatial-information} whereby all focal trees are assigned to one of `r n_fold` spatially contiguous blocks that act as folds in our cross-validation routine. Figure \ref{fig:scbi-spatial-cross-validation-schematic} presents a schematic illustrating this scheme for fold 1 as the test set and folds 2, 3, and 4 as the training sets. We fit the model to all focal trees in the training set, apply the model to all focal trees in the test set to compute fitted/predicted values, and compute the RMSE of the observed versus predicted growths. We repeat this procedure 3 more times with each of the three remaining folds acting as the test set. Furthermore, in order to maintain spatial independence between the test and training set, a buffer that extend outwards from the boundary of the test set is computed; all trees falling within this buffer will not be part of the training set.

```{r scbi-spatial-cross-validation-schematic, out.width="66%", fig.cap= "Schematic of spatial cross-validation: Using the k = 1 fold as the test set, assigning each focal tree to training set, test set, and buffer.", echo = FALSE}
fold_number <- 1

test_fold_buffer <-  blocks_scbi %>% 
  filter(folds == fold_number)

test_vs_train <- focal_vs_comp_scbi %>% 
  st_as_sf() %>% 
  add_buffer_variable(
    direction = "out", size = comp_dist, region = test_fold_buffer
  ) %>% 
  mutate(
    fold = case_when(
      foldID == fold_number ~ "test",
      foldID != fold_number & buffer ~ "training",
      foldID != fold_number & !buffer ~ "buffer"
    ),
    fold = factor(fold, levels = c("test", "buffer", "training"))
  )

ggplot() +
  geom_sf(data = test_fold_buffer, col = "red", fill = "transparent") +
  geom_sf(data = blocks_scbi, fill = "transparent", linetype = "dashed") +
  geom_sf_text(data = test_vs_train %>% st_as_sf() %>% sample_n(1000), aes(label = foldID, col = fold))
```

This algorithm is implemented in the `run_cv()` function, which acts as a wrapper function to the `comp_bayes_lm()` function that fits the model and the `predict()` generic that returns fitted values.

@@

```{r}
model_rmse_cv <- focal_vs_comp_scbi %>%
  rmse(truth = growth, estimate = growth_hat) %>%
  pull(.estimate)
model_rmse_cv
```

The resulting RMSE of `r model_rmse_cv` computed using cross-validation is `r ifelse(model_rmse_cv > model_rmse, "larger", "smaller")` than the earlier RMSE of `r model_rmse`, suggesting that models that do not take the inherent spatial autocorrelation of the data into account generate error estimates that are overly optimistic; in our case RMSE's that are too low.

@@

- show image of PLOSOne
- run time considerations
